Efficient pedigree recording for fast population genetics simulation

### 2018-01-12 forwards_paper.tex
Changed function notation slightly.

\include{macros}
\newcommand{\simupop}{\texttt{simuPOP}}
\newcommand{\fwdpp}{\texttt{fwdpp}}
\newcommand{\fwdpy}{\texttt{fwdpy11}}
\newcommand{\justc}{\texttt{C}}
\newcommand{\cpp}{\texttt{C++}}
\newcommand{\msprime}{\texttt{msprime}}
\newcommand{\ftprime}{\texttt{ftprime}}
\newcommand{\nodes}{\texttt{nodes}}
\newcommand{\edgesets}{\texttt{edgesets}}
\newcommand{\sites}{\texttt{sites}}
\newcommand{\mutations}{\texttt{mutations}}
\newcommand{\setdiff}{\smallsetminus}
\newcommand{\Nt}{\mathcal{N}}  % node table
\newcommand{\Et}{\mathcal{E}}  % edge table
\newcommand{\St}{\mathcal{S}}  % site table
\newcommand{\Mt}{\mathcal{M}}  % mutation table
\newcommand{\Al}{\mathcal{A}}  % list of ancestors
\newcommand{\priorityq}[1]{\mbox{PriorityQueue}({#1})}
\newcommand{\listnew}[1]{\mbox{List}(#1)}
\newcommand{\pqpush}[2]{#1.\mbox{push}(#2)}
\newcommand{\pqpop}[1]{#1.\mbox{pop}()}
\newcommand{\listappend}[2]{#1.\mbox{append}(#2)}
\newcommand{\ancsegment}[1]{\mbox{Segment}(#1)}
\newcommand{\taddrow}[2]{#1.\mbox{addrow}(#2)}
\newcommand{\attrparent}[1]{#1.\mbox{parent}}
\newcommand{\attrchild}[1]{#1.\mbox{child}}
\newcommand{\attrleft}[1]{#1.\mbox{left}}
\newcommand{\attrright}[1]{#1.\mbox{right}}
\newcommand{\attrid}[1]{#1.\mbox{id}}
% mbox to avoid hyphenating
\newcommand{\taddnode}[1]{\mbox{\texttt{addnode}}(#1)}
\newcommand{\taddedge}[1]{\mbox{\texttt{addedge}}(#1)}
\newcommand{\tsimplify}[1]{\mbox{\texttt{simplify}}(#1)}
\newcommand{\tsort}[1]{\mbox{\texttt{sort}}(#1)}
\newcommand{\tnodetable}[1]{\mbox{\texttt{NodeTable}}(#1)}
\newcommand{\tedgetable}[1]{\mbox{\texttt{EdgeTable}}(#1)}
\DeclareMathOperator{\len}{length}
\DeclareMathOperator{\add}{add} % for adding to tables
\newcommand{\lleft}{\mathrm{left}}  % left label
\newcommand{\lright}{\mathrm{right}}
\newcommand{\lid}{\mathrm{id}}

@@

We use $\randomuniform(A)$ to denote an element of the set $A$ chosen uniformly at random
(and all such instances are independent).
Given a node table $\Nt$, the function $\taddnode{\Nt, t}$
adds a new node to the table $\Nt$ with time $t$
and returns the ID of this new node.
Similarly, the function $\taddedge{\Et, \ell, r, p, c}$
adds a new edge $(\ell\mathrm{eft}, r\mathrm{ight}, p\mathrm{arent}, c\mathrm{hild})$ to the edge table $\Et$.
The function $\tsimplify{P, \Nt, \Et}$ (described below) % in section \ref{ss:simplify}
simplifies the history stored in the tables $\Nt$ and $\Et$
to the minimal information required to represent the genealogies of the list of node IDs $P$;
after simplification the nodes appearing in $P$ are relabeled $(0, 1, \ldots, |P|)$.
A step-by-step explanation follows the pseudocode.

@@

\algstep{W1.}{Initialisation.}{
    Set $\Nt \leftarrow \tnodetable{}$, $\mathcal{E}
    \leftarrow \tedgetable{}$, $t \leftarrow T$, and $j \leftarrow 0$.
 For $0 \leq k < N$, set $P_k \leftarrow \taddnode{\Nt, T}$.
}

\algstep{W2.}{Generation loop head: new node.}{Set $u \leftarrow \taddnode{\Nt, t}$ and $P'_j \leftarrow u$.
}

@@

\algstep{W4.}{Record edges.}{
Call $\taddedge{\Et, 0, x, P_a, u}$ and $\taddedge{\Et, x, L, P_b, u}$.
}

@@

The algorithm steps back up through the pedigree,
since nodes are ordered by time-since-birth.
First, for each input node $p$, we must find all ancestral segments inheriting from $p$ --
so, for every edge $(\ell, r, p, c)$,
Algorithm \algref{R} replaces any ancestry segment $[u, v)$ held by $c$
with $[u,v) \setdiff [\ell, r)$
(deleting the segment or splitting in two if necessary),
and adding the removed segment $[u,v) \cap [\ell, r)$ to the merge queue $Q$:

Next we must merge the ancestry segments in $Q$,
creating the new ancestor corresponding to input node $p$,
adding a node and edges to the output if any segments overlap,
which implies that an output coalescence has occurred.
Below we index elements of an ancestral segment $z = (\ell, r, u)$
by writing $z.\lleft = \ell$ and $z.\lright = r$ for the endpoints,
and $z.\lid = u$ for the output node ID.

>>>>

\include{macros}
\newcommand{\simupop}{\texttt{simuPOP}}
\newcommand{\fwdpp}{\texttt{fwdpp}}
\newcommand{\fwdpy}{\texttt{fwdpy11}}
\newcommand{\justc}{\texttt{C}}
\newcommand{\cpp}{\texttt{C++}}
\newcommand{\msprime}{\texttt{msprime}}
\newcommand{\ftprime}{\texttt{ftprime}}
\newcommand{\nodes}{\texttt{nodes}}
\newcommand{\edgesets}{\texttt{edgesets}}
\newcommand{\sites}{\texttt{sites}}
\newcommand{\mutations}{\texttt{mutations}}
\newcommand{\setdiff}{\smallsetminus}
\newcommand{\Nt}{\mathcal{N}}  % node table
\newcommand{\Et}{\mathcal{E}}  % edge table
\newcommand{\St}{\mathcal{S}}  % site table
\newcommand{\Mt}{\mathcal{M}}  % mutation table
\newcommand{\Al}{\mathcal{A}}  % list of ancestors
\newcommand{\priorityq}[1]{\mbox{\textbf{PriorityQueue}}({#1})}
\newcommand{\listnew}[1]{\mbox{\textbf{List}}(#1)}
\newcommand{\pqpush}[2]{#1.\mbox{\textbf{push}}(#2)}
\newcommand{\pqpop}[1]{#1.\mbox{\textbf{pop}}()}
\newcommand{\listappend}[2]{#1.\mbox{\textbf{append}}(#2)}
\newcommand{\ancsegment}[1]{\mbox{\textbf{Segment}}(#1)}
\newcommand{\taddrow}[2]{#1.\mbox{\textbf{addrow}}(#2)}
\newcommand{\attrparent}[1]{#1.\mbox{parent}}
\newcommand{\attrchild}[1]{#1.\mbox{child}}
\newcommand{\attrleft}[1]{#1.\mbox{left}}
\newcommand{\attrright}[1]{#1.\mbox{right}}
\newcommand{\attrid}[1]{#1.\mbox{id}}
% mbox to avoid hyphenating
\newcommand{\tsimplify}[1]{\mbox{\textbf{simplify}}(#1)}
\newcommand{\tsort}[1]{\mbox{\textbf{sort}}(#1)}
\newcommand{\tnodetable}[1]{\mbox{\textbf{NodeTable}}(#1)}
\newcommand{\tedgetable}[1]{\mbox{\textbf{EdgeTable}}(#1)}

@@

We use $\randomuniform(A)$ to denote an element of the set $A$ chosen uniformly at random
(and all such instances are independent).
Given a node table $\Nt$, the function $\taddrow{\Nt}{t}$
adds a new node to the table $\Nt$ with time $t$
and returns the ID of this new node.
Similarly, the function $\taddrow{\Et}{\ell, r, p, c}$
adds a new edge $(\ell\mbox{eft}, r\mathrm{ight}, p\mathrm{arent}, c\mathrm{hild})$ to the edge table $\Et$.
The function $\tsimplify{P, \Nt, \Et}$ (described below) % in section \ref{ss:simplify}
simplifies the history stored in the tables $\Nt$ and $\Et$
to the minimal information required to represent the genealogies of the list of node IDs $P$;
after simplification the nodes appearing in $P$ are relabeled $(0, 1, \ldots, |P|)$.
A step-by-step explanation follows the pseudocode.

@@

\algstep{W1.}{Initialisation.}{
    Set $\Nt \leftarrow \tnodetable{}$, $\mathcal{E}
    \leftarrow \tedgetable{}$, $t \leftarrow T$, and $j \leftarrow 0$.
 For $0 \leq k < N$, set $P_k \leftarrow \taddrow{\Nt}{T}$.
}

\algstep{W2.}{Generation loop head: new node.}{Set $u \leftarrow \taddrow{\Nt}{t}$ and $P'_j \leftarrow u$.
}

@@

\algstep{W4.}{Record edges.}{
Call $\taddrow{\Et}{0, x, P_a, u}$ and $\taddrow{\Et}{x, L, P_b, u}$.
}

@@

% The algorithm steps back up through the pedigree,
% since nodes are ordered by time-since-birth.
% First, for each input node $p$, we must find all ancestral segments inheriting from $p$ --
% so, for every edge $(\ell, r, p, c)$,
% Algorithm \algref{R} replaces any ancestry segment $[u, v)$ held by $c$
% with $[u,v) \setdiff [\ell, r)$
% (deleting the segment or splitting in two if necessary),
% and adding the removed segment $[u,v) \cap [\ell, r)$ to the merge queue $Q$:

% Next we must merge the ancestry segments in $Q$,
% creating the new ancestor corresponding to input node $p$,
% adding a node and edges to the output if any segments overlap,
% which implies that an output coalescence has occurred.
% Below we index elements of an ancestral segment $z = (\ell, r, u)$
% by writing $z.\lleft = \ell$ and $z.\lright = r$ for the endpoints,
% and $z.\lid = u$ for the output node ID.
